The Power of Unbiased Recursive Partitioning:\\ A Unifying View of CTree, MOB, and GUIDE

### 2019-06-05 pkg/partykit/inst/guideliketest/draft/draft.Rnw
revised discussion in Section 6
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2609 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 2104dba9c547e39a2f7bcbc087e3119bb8390278

\section{Discussion}
\label{sec:discussion}
From comparing the effects of the building blocks three main conclusions can be drawn: 
(1)~In many cases including all available score values rather than just residuals leads to 
clearly better results and performs in general at least equally well.
(2)~Dichotomizing scores/residuals does not lead to any improvement in performance.
(3)~Categorizing possible split variables also doesn't improve the performance 
unless the breaks used for the categorization are very close to the true split points.
However, even then the results are very similar to those achieved without any categorization.
But it has to be noted that a categorization of possible split variables can simplify
evaluations. Moreover, depending on the focus of interest, the results can be more
straightforward, for example in case of clearly separated subgroups being defined by the values 
of a continuous split variable. Therefore, categorizing possible split variables can have
advantages in practice, however, it has to be considered that in general
important effects might be neglected.

Both of the transformation functions $h$ and $g$ (applied to the response~$Y$ or a split 
variable~$Z$ respectively) can also be multivariate. For this case, the corresponding
statistical tests are available, in particular in the \textsf{R} package \textbf{coin}
which is employed for the implementation of MOB and CTree and provides 
the required toolbox for multivariate dependency tests, including a $\chi^2$~test.

Linear models as presented in this study have been selected as they allow for a simple illustration 
of the investigated subjects. However, the obtained results also hold for a variety of other 
models as the introduced building blocks can be applied in the same way for all model types.
Specific characteristics of linear models regarding structural changes have been investigated
in detail in previous literature.
\cite{Ploberger+Kraemer:1992} have shown that residual-based tests can detect a change in the parameters
only if it also causes a shift in the expected value $\mathbb{E}(Y)$. 
This is not the case if changes are orthogonal to the mean regressor $(1,0)^{\top}$, hence, if
the parameter vector $\beta$ is shifted 
%to $\beta + \Delta$ 
by $\Delta = (0, \Delta_1)^{\top}$.
For that reason GUIDE breaks down when the slope parameter $\beta_1$ changes while the intercept
$\beta_0$ is fixed as seen in the middle panel of Figure~\ref{fig:scores_stump}.
While this might be an extreme example similar scenarios can appear in many applications, see, e.g., 
Figure~2 in \cite{Loh+He+Man:2015} where an additional separation of the data in each node is
required in order to fit a reasonable model. This can be avoided by including model scores
allowing for the detection of changes not only in the location parameter but also in all other 
parameters, i.e., not only in the expected value but also in the variance.
In situations where shifts appear only in the intercept testing procedures purely based on residuals 
perform best. However, the cost in terms of power that comes with including all scores, in 
particular with testing more than one parameter, is negligibly small.

Also in more general models residual-based and score-based procedures are expected to perform
equally well if all parameters are highly correlated. On the other hand, if parameters are 
orthogonal changes that are not connected to the mean might be missed when only considering
residuals. This can be prevented by including scores which are a standard measurement of goodness 
of fit that can be easily employed for all model types. Moreover, the score function~$s$ can 
simply be seen as a transformation to a different space in order to allow for a well structured
analysis of dependencies, regardless of the type of model and therefore applicable in a
wide variety of situations.
In order to benefit from the advantages of using scores 
some already existing models have been adapted, e.g., in \cite{Hothorn+Zeileis:2017} 
and \cite{Schlosser+Hothorn+Stauffer:2019}. Similarly, variations of the GUIDE algorithm
using different types of residuals are provided, e.g., in \citealp{Loh+Zheng:2013}, 
\citealp{Chaudhuri+Loh:2002} and \citealp{Loh+He+Man:2015}.
But even though it also depends on the type of the applied residuals whether parameter changes 
can be detected successfully, a one-dimensional measurement of goodness of fit such as residuals
provides less power than full model scores. 
Whether this difference affects the performance of a testing strategy and if so, to what extent, is 
determined by the way how residuals are connected to parameter shifts as well as the type of parameter
shift and the type of model. Therefore, a general statement for all types of models and residuals can 
hardly be made. However, the results from previous literature such as \cite{Ploberger+Kraemer:1992} and
the here presented simulation show that considering only residuals provides small gains in certain 
situations but remarkable deficits in general. On the contrary, the benefits of including full model 
scores clearly outweigh the price of testing all parameters.

%Furthermore, with only two parameters determining a linear model the importance of including scores 
%regarding each parameter could already be pointed out clearly while for an increasing number of 
%parameters in a model this aspect becomes even more important.

\section*{Computational details}
\label{sec:comp}
The applied implementation is based on the \textsf{R} package \textbf{partykit}
(version~\Sexpr{packageVersion("partykit")}) which is available 
on \textsf{R}-Forge at (\url{https://R-Forge.R-project.org/projects/partykit/})
where the code to reproduce the simulation study can be found as well.

@@

\section{Results over increasing effect size}
\label{app:increasingdelta}

>>>>

\section{Discussion}
\label{sec:discussion}

The testing strategies underlying the unbiased recursive partitioning algorithms
CTree, MOB, and GUIDE have been embedded in a common inference framework, highlighting
what the tests have in common and what sets them apart. Concerning the effects 
of the corresponding building blocks for the tests, three main conclusions can be drawn: 
%
\begin{itemize}
  \item \emph{Goodness-of-fit measure:} Assessing all dimensions of a model via the full
        scores is to be preferred over assessing only a subset with the residuals. The former
	can substantially improve performance while leading only to minor deteriorations
	when it is not necessary.
  \item \emph{Dichotomization of residuals/scores:} No scenarios could be found where this
        is beneficial and tests without dichotomization performed clearly better in several
	scenarios.
  \item \emph{Categorization of split variables:} The effects of categorization are not so
        clear-cut and depend on the true data structure. If there are indeed abrupt shifts
	close to the breaks from the categorization, it works well. However, for splits
	close to the margins performance can deteriorate and a maximally-selected test is
	preferable. Finally, a linear statistic performs better for smooth rather than
	abrupt changes.
\end{itemize}
%
Thus, also when categorization as in GUIDE is used we would recommend to employ the non-dichotomized
scores instead of the dichotomized residuals. Note that such a test corresponds to a
multiple ANOVAs (for the score components) and is easily available in statistical software.
Moreover, in the \proglang{R} system the \pkg{coin} package provides a convenient toolbox that
encompasses multivariate transformation functions $h$ and/or $g$.

Linear models as presented in this study have been chosen as they are highly relevant in practice,
allow for simple illustrations, and theoretical insights are available for the testing strategies.
However, the results can be easily extended to a wide variety of other models where the introduced building
blocks can be applied in the same way. Specifically, it has been shown theoretically that certain
some changes in the parameters do not lead to shifts in the residuals (but in other components of the
scores). \cite{Ploberger+Kraemer:1992} showed that residual-based tests can detect a change in the parameters
of a linear model only if it also causes a shift in the expected value $\mathbb{E}(Y)$. 
This is not the case if changes are orthogonal to the mean regressor which in our case is $(1,0)^{\top}$.
Consequently, if only the slope $\beta_1$ but not the intercept $\beta_0$ changes, the shift
is of type $(0, \delta)^\top$ and thus orhtogonal to the mean regressor. Due to this residual-based
tests as in GUIDE break down and do not have power to detect this. Note that this situation does not
have to be rare in practice: Especially for binary regressor variables (e.g., as in treatment-subgroup
investigations) it can easily occur \citep[see Figure~2 in][for an illustration]{Loh+He+Man:2015}.

Also in more general models, residual-based and score-based procedures are expected to perform
equally well if all model parameters are highly correlated. But if parameters do change orthogonally
this might again be missed when only considering residuals -- and full model scores are typically
easily available as the appropriate remedy. Note that the score function~$s$ can also
simply be seen as a transformation of the response variable $Y$ (and potentially regressors $X$)
to a different space in order to allow for a well structured analysis of dependencies. This has
been exploited in several tree-based approaches previously published in the literature,
e.g., in \cite{Hothorn+Zeileis:2017} and \cite{Schlosser+Hothorn+Stauffer:2019}. Similarly, it would
be of interest to investigate a score-based version of the extended GUIDE algorithms beyond the
linear model, e.g., in \citealp{Loh+Zheng:2013}, \citealp{Chaudhuri+Loh:2002}, and \citealp{Loh+He+Man:2015}.

\section*{Computational details}
\label{sec:comp}
The applied implementation is based on the \proglang{R} package \pkg{partykit}
(version~\Sexpr{packageVersion("partykit")}) which is available 
on \proglang{R}-Forge at (\url{https://R-Forge.R-project.org/projects/partykit/})
where the code to reproduce the simulation study can be found as well.

@@

\section{Results for increasing effect size}
\label{app:increasingdelta}
